---
title: "Willamette Falls Fish Passage Time Series: A Summary"
author: "Travis Renacker"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    number_sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show Code"
    embed-resources: true
    theme: superhero
execute:
  eval: true
  message: false
  warning: false


---


# Overview

Overview (above the tabs) should contain, in whatever order you choose: 
•	An overview section with subsections that briefly summarize the dataset (this should include a well formatted data citation - provider, title, URL, date accessed - use info and links above to find all this out), the purpose of your analysis, and a pseudocode outline of the steps of your analysis.  Your writeup should look and feel professional in style, tone, and substance.

•	An engaging image (with caption, incl. photo credit) that is relevant to the dataset

•	Your organized code, with clear subsections and any useful descriptive text / annotation (e.g. if you wanted to highlight this as a code example for a prospective employer). 

•	All code, including attached packages, should be included using code-folding.  Make sure to suppress any messages & warnings. Set embed-resources to be true so your HTML is self-contained!

# Library

```{r setup}
library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library(feasts)
library(fable)
library(patchwork)
library(ggstream)
```



#  Salmon in Willamette Passage

#### Raw Data

```{r}

# Load the raw data

fish_data_raw <- read_csv(here("data", "willamette_fish_passage.csv")) |> janitor::clean_names()

```



## Data Wrangle and Tsibble Creation

#### Coho, Jack Coho, and Steelhead

```{r}

#filter for three species, replace NA with 0, and pivot longer columns "fish_type" and "count" and mutate date into date format instead of character

three_salmon <- fish_data_raw |> 
  select(date, coho, jack_coho, steelhead) |>  
  mutate(across(c(coho, jack_coho, steelhead), ~ replace_na(., 0))) |>  #NA to 0
  pivot_longer(
    cols = c(coho, jack_coho, steelhead),  # Columns to pivot
    names_to = "fish_type",  # column to hold the names of the fish types
    values_to = "count"  # column to hold the fish count
  ) |> 
  mutate(date = mdy(date))  # Convert character 'date' to Date format

```



```{r}
#| output: false

#data vizualization and exploration

ggplot(data = three_salmon) +
  geom_point(aes(x = date, y = count, color = fish_type)) + 
  scale_color_brewer(palette = "Dark2")
```
Create a tsibble 
```{r}

#monthly stibble 
salmons_ts <- three_salmon |> 
  # Create a month/year formatted as character
  mutate(month_year_char = format(date, "%m/%Y"),
         # Also create the yearmonth object for the tsibble
         month_year = yearmonth(date)) |> 
  # Group by month/year and fish_type
  group_by(month_year, fish_type) |> 
  summarise(monthly_count = sum(count), 
            month_year_char = first(month_year_char), 
            .groups = "drop") |> 
  # Create the tsibble
  as_tsibble(key = fish_type, index = month_year)

#daily stibble for salmon 
salmon_daily <- three_salmon |> 
  mutate(date = as.Date(date)) |> 
  group_by(date, fish_type) |> 
  summarise(daily_count = sum(count), .groups = "drop") |> 
  as_tsibble(key = fish_type, index = date)

```



## Original Time Series

### Coho, Jack Coho, and Steelhead

```{r} 

# For the monthly time series plot
Monthly <- ggplot(data = salmons_ts, aes(x = month_year, y = monthly_count, colour = fish_type)) +
  geom_line() + 
  scale_colour_brewer(palette = "Dark2", labels = c("Coho", "Jack Coho", "Steelhead")) +  # Use Dark2 color palette
  labs(
    x = "Date",
    y = "Salmon Count at Willamette Falls",
    title = "Monthly Salmon Counts by Fish Type"
  ) + 
  theme_minimal() +  # Use a minimal theme for the plot
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for better readability
  guides(colour = guide_legend(title = "Salmon Type", 
                               title.position = "top", 
                               title.hjust = 0.5))

# Daily Time Series Plot
Daily <- ggplot(data = salmon_daily, aes(x = date, y = daily_count, colour = fish_type)) +
  geom_line() + 
  scale_colour_brewer(palette = "Dark2", labels = c("Coho", "Jack Coho", "Steelhead")) +   # Use Dark2 color palette
  labs(
    x = NULL,
    y = "Salmon Count at Willamette Falls",
    title = "Daily Salmon Counts by Fish Type"
  ) + 
  theme_minimal() +  # Use a minimal theme for the plot
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for better readability
  guides(colour = guide_legend(title = "Salmon Type", 
                               title.position = "top", 
                               title.hjust = 0.5))
                                 
```

::: {.panel-tabset}

## Daily Time Series

```{r}
#| fig-cap: "**Figure 1**. Daily Salmon Counts by Fish Type from 2001-2010. Coho, Jack Coho and Steelhead shown."

Daily 
```


## Monthly Agraggated Time Series

```{r}
#| fig-cap: "**Figure 2**. Monthly Salmon Counts by Fish Type from 2001-2010. Coho, Jack Coho and Steelhead shown."

Monthly

```


## Notable Trends 

### Coho, Jack Coho, and Steelhead Salmon in Willamette Passage 2000-2010

 -  During this scale of time there appears to be a clear seasonality for each species. The time scale appears to be too narrow to parse out cyclical trends. 
 
 - The Coho and Jack Coho appear to have similar trends, with the Jack Coho having a slightly lower count. Both species appear to be increasing multiplicatively. 
 
 - The Steelhead appear to have a flat, to possibly slight negative trend.
 
 - Both Coho species appear to have a seasonal peak in September while the Steelhead appear to have a seasonal peak around May. 

########################################################################################
 Does there appear to be an overall trend?
-   Does there appear to be seasonality?
-   Does there appear to be cyclicality?
-   Any notable outliers or additional patterns?

#######################################################################################

:::


```{r}
#coho plot 
coho_p <- salmon_daily %>%
  filter(fish_type == 'coho') %>%
  gg_season(daily_count, pal = rev(viridis::plasma(n = 10))) + 
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "cornsilk", fill = NA)
  ) +
  labs(
    x = NULL,
    y = "Salmon Count",
    title = "Coho",
    color = "Year"  # This changes the legend title
  )
  
#jack coho plot
jack_p <- salmon_daily %>%
  filter(fish_type == 'jack_coho') %>%
  gg_season(daily_count,  pal = rev(viridis::plasma(n = 10)))  + 
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "cornsilk", fill = NA)
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Jack Coho",
    color = "Year"  # This changes the legend title
  )

# Enhanced steelhead plot
steelhead_p <- salmon_daily %>%
  filter(fish_type == 'steelhead') %>%
  gg_season(daily_count, pal = rev(viridis::plasma(n = 10))) + 
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "cornsilk", fill = NA)
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Steelhead",
    color = "Year",
    caption = "Data source: Columbia Basin Research"
  )

```


```{r}
#| output: false

# Remove all individual legends
jack_plot <- jack_p + theme(legend.position = "none")
coho_plot <- coho_p + theme(legend.position = "none")
steelhead_plot <- steelhead_p + theme(legend.position = "none")

# Combine with a collected legend on the right
combined_plots <- (jack_plot / coho_plot / steelhead_plot) + 
  plot_layout(heights = c(1, 1, 1)) +   # Stack plots vertically with equal height
  plot_layout(guides = "collect") +  # Collect all legends into one
  theme(legend.position = "right",   # Place the collected legend on the right
        plot.margin = margin(10, 10, 10, 10)) +  # Adjust plot margins
  plot_annotation(title = "Salmon Passage by Species")  # Add a title for the combined plot

# Display the combined plot
combined_plots

```

## SeasonPlots

::: {.panel-tabset}

## Steelhead

```{r}
#| fig-cap: "**Figure 3**. Steelhead Seasonality from 2001-2010 at Wllamette fish passage."

# Enhanced steelhead plot
 salmon_daily %>%
  filter(fish_type == 'steelhead') %>%
  gg_season(daily_count, pal = rev(viridis::plasma(n = 10))) + 
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "cornsilk", fill = NA)) +
  labs(
    x = NULL,
     y = "Salmon Count",
    title = "Steelhead",
    color = "Year",
    caption = "Data source: Columbia Basin Research")
 
```

## Jack Coho

```{r}
#| fig-cap: "**Figure 4**. Jack Coho Seasonality from 2001-2010 at Wllamette fish passage."

#jack coho plot
salmon_daily %>%
  filter(fish_type == 'jack_coho') %>%
  gg_season(daily_count,  pal = rev(viridis::plasma(n = 10)))  + 
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "cornsilk", fill = NA)) +
  labs(
    x = NULL,
     y = "Salmon Count",
    title = "Jack Coho",
    color = "Year",
    caption = "Data source: Columbia Basin Research")
```

## Coho

```{r}
#| fig.cap: "**Figure 5**. Coho Seasonality from 2001-2010 at Wllamette fish passage."

#coho plot 
salmon_daily %>%
  filter(fish_type == 'coho') %>%
  gg_season(daily_count, pal = rev(viridis::plasma(n = 10))) + 
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "cornsilk", fill = NA)) +
  labs(
    x = NULL,
    y = "Salmon Count",
    title = "Coho",
    color = "Year", 
    caption = "Data source: Columbia Basin Research")

```


## Notable Trends 

### Seasonplot for Three Salmon Species

- By overlaying each years data signified with a different color onto a single plot, it becomes easier to assess the seasonality for each species. 

- Both Coho and Jack Coho have a similar seasonal pattern, with a tight clustering around the peak in October.

- Steelhead, on the other hand, have a peak over the summer. The seasonality is not as clearly defined as with the other two, showing peaks in multiple months across the time series. 

:::


## Annual Counts

::: {.panel-tabset}

## Annual Counts

```{r}
#| fig.cap: "**Figure 6**. Yearly Fish Totals for Three Fish Species from 2001-2010 in Wllamette fish passage."

salmon_daily %>% 
  mutate(year = year(date)) %>%  # Pull year from date
  ggplot(aes(x = year, y = daily_count, fill = fish_type)) +
  geom_bar(stat = 'identity', position = 'stack') +  
  scale_fill_brewer(palette = "Dark2", labels = c("Coho", "Jack Coho", "Steelhead")) +  # Color palette
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+  # Ensure whole number labels
  labs(x = NULL, y = 'Count', fill = 'Salmon Species', caption = "Data source: Columbia Basin Research") +  
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "cornsilk", fill = NA),
    panel.background = element_rect(fill = "white")  # Optional: to ensure a white background
  )

```
## Proportional Annual Counts

```{r}
#| fig.cap: "**Figure 7**. Proportional Yearly Fish Totals for Three Fish Species from 2001-2010 in Wllamette fish passage."

proportional_salmon <- three_salmon %>%
  mutate(year = year(date)) %>%
  group_by(year, fish_type) %>%
  summarize(count_sum = sum(count, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  mutate(proportion = count_sum / sum(count_sum)) %>%
  ungroup()

ggplot(proportional_salmon, aes(x = year, y = proportion, fill = fish_type)) +
  geom_stream(type = "proportional") +
  scale_fill_brewer(palette = "Dark2", labels = c("Coho", "Jack Coho", "Steelhead")) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportional Distribution of Three Salmon Types",
    x = NULL,
    y = NULL,
    fill = NULL, 
    caption = "Data source: Columbia Basin Research"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

```
## Notable Trends 

### Major Yearly Trends


- From the yearly totals it is clear that there has been a drastic increase in the number of Coho Salmon. 


- The total number of fish observed each year has stayed somewhat flat. 


- As Coho and Jack Coho increase in occurrence the Steelhead appear to diminish within this time series.  





:::

Optional: Challenge yourself on task 2:
•	Change the Headers into Tabs: Tabs are interactive mini-pages you can add to your document to better organize your results. Use the Quarto documentation page to put each section on a separate tab.
•	Forecast Salmon runs with Holt-Winters: Apply the Holt-Winters forecast method on the original time series data for each salmon population. What future trends does the model predict? How applicable is this forecast method for salmon? Show your results in a fourth section/tab.
